### 2024-04-07 paper/main.tex
paper

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Mamba-Based Tumor Segmentation\\
\thanks{$\dagger$ Equal contributions}
}

\author{\IEEEauthorblockN{Clay Crews$^{\dagger}$}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of South Carolina}\\
Columbia, SC, United States \\
jccrews@email.sc.edu}
\and
\IEEEauthorblockN{Lexington Whalen$^{\dagger}$}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of South Carolina}\\
Columbia, SC, United States\\
LAWHALEN@email.sc.edu}
}

\maketitle

\begin{abstract}
U-net and transformer-based architectures have dominated medical image segmentation, particularly for tumor segmentation from brain MRI scans. However, Structured State Space Sequence Models (SSMs), as highlighted by Gu, Goel, and Ré, present a promising alternative with comparable modeling power to transformers but linear scalability in sequence length, making them well-suited for large medical images.

We propose exploring the application of the Mamba block, a variant of SSMs, for tumor segmentation in brain MRI scans. By leveraging selective state spaces, Mamba blocks can potentially enhance the efficiency and effectiveness of tumor segmentation while maintaining linear scalability.

Our methodology involves implementing and fine-tuning a Mamba-based architecture on a dataset of annotated brain MRI scans. We will evaluate the model's performance using standard segmentation metrics and compare the results with U-net and transformer-based approaches.
\end{abstract}

\begin{IEEEkeywords}
Mamba, Segmentation, Selective State Spaces
\end{IEEEkeywords}

\section{Introduction}
MRI imaging to identify brain tumors has naturally been a target for accurate image segmentation through the use of ML architecture. The use of computer-aided diagnosis of tumors can be crucial to quickly identify tumors and determine a course of action for the patient. Gliomas are the most common primary brain malignancy with varying degrees of aggressiveness in the brain. The protocol for MRI image annotation of these tumors consists of the following labels: the whole tumor extent, the tumor core, the non-enhancing/necrotic tumor region, and regions of low grade gliomas \cite{bakas2019identifying}. These annotations of an MRI are used to determine the size and severity of a tumor.

To accurately segment these images, picking out the small features of the image becomes the focus. The use of the U-Net architecture, presented by Ronneberger, Fischer, and Brox \cite{ronneberger2015unet}, has had success in MRI image segmentation. U-Net builds on the approach of fully convolutional networks by improving the limitation on the large amount of data needed to accurately train the model. Very few training images are needed due to the contracting and expansive paths in this architecture. In the contracting path, images are downsampled and pooled along with increasing the number of feature channels. Each downsampled resolution produces a multi-channel feature map. The expansive path up samples and combines with each segmentation feature map from the contracting layer to localize features in the image. A 1x1 convolution layer is applied to the final image to map the large feature vector to a selected number of classes. All of this essentially creates a feature map where each pixel’s relevance is taken into account and evaluated in the end result. An approach using the U-Net architecture would provide the detailed analysis needed for tumor annotation.

An arising issue in the complex recurrent or convolutional neural networks used for image segmentation is the length of the sequences given as input. Maintaining relevance of the current area of an image in relation to the rest of the image is key for more accurate segmentation. An approach with the Transformer architecture has a large emphasis on learning specific features in an image, outlined by Vaswani et al. 2017 \cite{vaswani2023attention}. This architecture follows an encoder-decoder pattern, connected through an attention mechanism, a transformer block. Transformer blocks show capability of learning long-distance dependency throughout an input image. In the self-attention mechanism for this block, a single element in a given sequence is compared to all of the elements in the sequence. However, this mechanism becomes computationally very expensive when it comes to long sequences.

Selective State Spaces (SSMs) pose an improvement in computational efficiency over Transformers for this long dependency range. SSMs represent the relevance and context of different parts of the input sequence giving selective attention. This model is widely used in control theory for time variant systems and in fields such as computational neuroscience. Selective attention is very efficient in handling long sequences over time and presents a more localized context of the data. Sections of sequences that deserve attention are represented by the model and will dynamically update these values to reflect the contextual relevance.

Making use of Structured State Spaces, presented by Gu, Goel, and Ré \cite{gu2022efficiently}, for their superiority in long range dependency tasks, a specific type of SSM, the Mamba architecture, by Gu and Dao \cite{gu2023mamba}, proposed a foundational model to operate on arbitrary sequences from a variety of inputs in the domain of sequence modeling. This model achieves the power of Transformers while scaling linearly in sequencing length to be computationally efficient. Additionally, the selection mechanism of Mamba improves on an SSMs ability to focus on or ignore sections in an input sequence by parameterizing the parameters of an SSM to reflect the current context of the input. The Mamba model and its application to brain tumor MRI imaging segmentation will be evaluated in this project.

\section{Compared Models}
Prior to an analysis of how the Pyramidal U-Mamba compares, we shall explain what models we chose for comparison and why.

For this our analysis, we compare four models against our's. The models are U-Net\cite{ronneberger2015unet},ResNet18, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}. Below we explain our choices.
\subsection{U-Net}
 The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information [1]. 
 A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. Unlike traditional convolutional networks, U-Net does not employ any fully connected layers and only utilizes the valid part of each convolution [2]. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{DBLP:journals/corr/abs-2006-11239} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) (Gu et al., 2022), a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegViT}
SegViT \cite{zhang2022segvit} explores the capability of plain Vision Transformers (ViTs) \cite{dosovitskiy2021image} for semantic segmentation, leveraging the attention mechanism to generate masks for semantic segmentation. Unlike previous ViT-based segmentation networks that learn pixel-level representations from the output of the ViT, SegViT introduces the Attention-to-Mask (ATM) module. In the ATM module, the similarity maps between a set of learnable class tokens and the spatial feature maps are transferred to the segmentation masks, enabling effective semantic segmentation.
SegViT has demonstrated impressive performance on several benchmark datasets, including ADE20K, COCO-Stuff-10K, and PASCAL-Context, achieving state-of-the-art results. Moreover, to address the computational cost associated with the ViT backbone, SegViT proposes query-based down-sampling (QD) and query-based up-sampling (QU) to build a Shrunk structure. By comparing U-Mamba against SegViT, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the attention-based segmentation mechanism employed by SegViT. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of attention mechanisms and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

\section{Results}
We now compare our model against three other common segmentation models, U-Mamba \cite{}, U-Net \cite{}, and SegViT \cite{}. And it was good.

\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
Model & Param Count & Size (MB) & Avg. IOU & Avg. Dice \\
\hline
Pyramidal Mamba & Y & Z & A & B \\
\hline
U-Mamba & Y & Z & A & B \\
\hline
U-Net & Y & Z & A & B \\
\hline
SegViT & Y & Z & A & B \\
\hline
\end{tabular}

\section{Loss Curves}
We also display the loss curves for each model. Each model was trained for 100 epochs under [OTHER CONDITIONS].

\section*{Acknowledgment}

The preferred spelling of the word ``acknowledgment'' in America is without 
an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
acknowledgments in the unnumbered footnote on the first page.

\begin{thebibliography}{00}
\bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.

\bibitem{vaswani2023attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2023.

\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.

\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.

\bibitem{bakas2019identifying} S. Bakas et al., "Identifying the Best Machine Learning Algorithms for Brain Tumor Segmentation, Progression Assessment, and Overall Survival Prediction in the BRATS Challenge," arXiv:1811.02629 [cs.CV], 2019.

\bibitem{U-Mamba} J. Ma, F. Li, and B. Wang, "U-Mamba: Enhancing Long-range Dependency for Biomedical Image Segmentation," arXiv:2401.04722 [cs.CV], 2024.

\bibitem{rombach2021highresolution} R. Rombach, A. Blattmann, D. Lorenz, P. Esser, and B. Ommer, "High-Resolution Image Synthesis with Latent Diffusion Models," arXiv:2112.10752 [cs.CV], 2021.

\bibitem{DBLP:journals/corr/abs-2006-11239} J. Ho, A. Jain, and P. Abbeel, "Denoising Diffusion Probabilistic Models," CoRR, vol. abs/2006.11239, 2020. [Online]. Available: https://arxiv.org/abs/2006.11239

\bibitem{DBLP:journals/corr/HeZRS15} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," CoRR, vol. abs/1512.03385, 2015. [Online]. Available: http://arxiv.org/abs/1512.03385

\bibitem{zhang2022segvit} B. Zhang et al., "SegViT: Semantic Segmentation with Plain Vision Transformers," arXiv:2210.05844 [cs.CV], 2022.

\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\end{thebibliography}
\end{document}

### 2024-04-09 tests/BCEDiceLoss.py
ultra lightm-unet results

@@ -1,0 +1,47 @@
+import torch
+import torch.nn as nn
+
+class BCELoss(nn.Module):
+    def __init__(self):
+        super(BCELoss, self).__init__()
+        self.bceloss = nn.BCELoss()
+
+    def forward(self, pred, target):
+        size = pred.size(0)
+        pred_ = pred.view(size, -1)
+        target_ = target.view(size, -1)
+
+        return self.bceloss(pred_, target_)
+
+
+class DiceLoss(nn.Module):
+    def __init__(self):
+        super(DiceLoss, self).__init__()
+
+    def forward(self, pred, target):
+        smooth = 1
+        size = pred.size(0)
+
+        pred_ = pred.view(size, -1)
+        target_ = target.view(size, -1)
+        intersection = pred_ * target_
+        dice_score = (2 * intersection.sum(1) + smooth)/(pred_.sum(1) + target_.sum(1) + smooth)
+        dice_loss = 1 - dice_score.sum()/size
+
+        return dice_loss
+
+
+class BceDiceLoss(nn.Module):
+    def __init__(self, wb=1, wd=1):
+        super(BceDiceLoss, self).__init__()
+        self.bce = BCELoss()
+        self.dice = DiceLoss()
+        self.wb = wb
+        self.wd = wd
+
+    def forward(self, pred, target):
+        bceloss = self.bce(pred, target)
+        diceloss = self.dice(pred, target)
+
+        loss = self.wd * diceloss + self.wb * bceloss
+        return loss

### 2024-04-19 paper/main.tex
updated results table
update table

@@
 
+\begin{table}[!h]
+    \renewcommand{\arraystretch}{1.2}
+    \setlength{\tabcolsep}{2pt}
     \centering
-\begin{tabular}{|c|c|c|c|c|}
+    \footnotesize
+    \begin{tabular}{|l|c|c|c|c|}
     \hline
-Model & Param Count & Size (MB) & Avg. IOU & Avg. Dice \\
+    Model & Params & Size & IOU & Dice \\
     \hline
-Pyramidal Mamba & Y & Z & A & B \\
+    Pyramidal Mamba (Ours) & 10M & 109.15 & 0.8004 & 0.8867 \\
     \hline
-U-Mamba & Y & Z & A & B \\
+    U-Mamba & 9.8M & 108.10 & 0.7908 & 0.8799 \\
     \hline
-U-Net & Y & Z & A & B \\
+    U-Net & 7.8M & 31.09 & 0.8071 & 0.8929 \\
     \hline
@@
     \hline
+    ResNet18 & 13.9M & 55.76 & 0.7623 & 0.8648 \\
+    \hline
+    ResNet50 & 42.9M & 171.85 & 0.7661 & 0.8672 \\
+    \hline
+    Ultra LightM-UNet & 0.049M & 0.20 & 0.6725 & 0.7955 \\
+    \hline
     \end{tabular}
+    \end{table}
 
-\section{Loss Curves}
-We also display the loss curves for each model. Each model was trained for 100 epochs under [OTHER CONDITIONS].
+\section{Acknowledgment}
 
-\section*{Acknowledgment}
-
-The preferred spelling of the word ``acknowledgment'' in America is without 
-an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
-G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
-acknowledgments in the unnumbered footnote on the first page.
+This work was done as a final project for the University of South Carolina's CSCE763 Digital Image Processing course taught in Spring 2024 under Dr. Yan Tong.
 

### 2024-04-20 paper/main.tex
paper

@@
 
+\subsection{UltraLight VM-UNet}
+The UltraLight VM-UNet \cite{ultralightvmunet} is a lightweight neural network architecture designed for skin lesion segmentation tasks.
+It is built upon the Vision Mamba module, which is a state-space model (SSM) that can efficiently handle long-range dependencies in sequences, making it well-suited for image segmentation tasks.
+
+The key innovation of the UltraLight VM-UNet is the proposed Parallel Vision Mamba Layer (PVM Layer), which processes deep features in parallel using multiple Vision Mamba blocks.
+Specifically, the input feature map is split into multiple sub-feature maps, each processed by a separate Vision Mamba block with a reduced channel count.
+This parallel processing approach allows the UltraLight VM-UNet to maintain high segmentation performance while significantly reducing the number of parameters and computational complexity.
+
+The UltraLight VM-UNet is reported to have only 0.049 million parameters and a computational cost of 0.060 GFLOPs, which is significantly lower than traditional convolutional neural networks and transformers used for image segmentation tasks.
+Despite its lightweight nature, the authors demonstrate that the UltraLight VM-UNet achieves competitive performance on three publicly available skin lesion segmentation datasets, outperforming several state-of-the-art lightweight models.
+
+The success of the UltraLight VM-UNet can be attributed to the authors' in-depth analysis of the key factors influencing the parameters of the Vision Mamba module.
+By identifying the number of input channels as a critical factor affecting the parameter count, they were able to design the PVM Layer to process features in parallel while keeping the overall channel count constant, leading to a significant reduction in parameters without compromising performance.
+
+\section{Our Models}
+
+Inspired by \cite{U-Mamba} and \cite{ultralightvmunet}, we sought to implement both models on a different problem: tumor segmentation.
+
+Our goal was to 1) maintain dice scores comparable to state-of-the-art (SOTA) methods such as those listed above, while 2) being smaller than those above.
+
+We now go into what we did for each of our models.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/bars.png}
+    \caption{Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
+    \label{fig:bars}
+\end{figure}
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/balls.png}
+    \caption{A scatter plot of Dice Score vs model size in MegaBytes. The points are scaled to help represent size.}
+    \label{fig:balls}
+\end{figure}
+
+\subsection{PP-UMamba}
+This model incorporates pyramidal pooling, which is a strategy used in convolutional neural networks (CNNs) to capture context and incorporate multi-scale information \cite{zhao2017pyramid}.
+Pyramidal pooling involves parallel pooling operations at different scales, followed by concatenation of the resulting feature maps.
+This approach has been shown to improve the performance of CNNs in various computer vision tasks, including segmentation, by enabling the model to capture both local and global context.
+
+In our PP-UMamba model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
+
+\subsection{ULMUNet\_v1}
+For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder.
+Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step.
+The intention was to improve performance by increasing the capacity of the model while maintaining a reasonable parameter count.
+
+\subsection{ULMUNet\_v2}
+In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8.
+This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.
+
+\subsection{ULMUNet\_v3}
+ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2.
+It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256].
+Additionally, the depth of the U-Net was reduced to 5 layers.
+This approach aims to balance the trade-off between model capacity and parameter efficiency.
+
+\subsection{ULMUNet\_v4}
+This model incorporates pyramidal pooling, similar to PP-UMamba, but within the UltraLight VM-UNet architecture.
+The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.
+
+\subsection{UL-PP}
+The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters.
+According to the provided information, incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+
 \section{Results}
-We now compare our model against three other common segmentation models, U-Mamba \cite{}, U-Net \cite{}, and SegViT \cite{}. And it was good.
+We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck, UNet, and UltraLight VM-UNet. The results are shown in Figures 1--2.
 
-\begin{table}[!h]
-    \renewcommand{\arraystretch}{1.2}
-    \setlength{\tabcolsep}{2pt}
-    \centering
-    \footnotesize
-    \begin{tabular}{|l|c|c|c|c|}
-    \hline
-    Model & Params & Size & IOU & Dice \\
-    \hline
-    Pyramidal Mamba (Ours) & 10M & 109.15 & 0.8004 & 0.8867 \\
-    \hline
-    U-Mamba & 9.8M & 108.10 & 0.7908 & 0.8799 \\
-    \hline
-    U-Net & 7.8M & 31.09 & 0.8071 & 0.8929 \\
-    \hline
-    SegViT & Y & Z & A & B \\
-    \hline
-    ResNet18 & 13.9M & 55.76 & 0.7623 & 0.8648 \\
-    \hline
-    ResNet50 & 42.9M & 171.85 & 0.7661 & 0.8672 \\
-    \hline
-    Ultra LightM-UNet & 0.049M & 0.20 & 0.6725 & 0.7955 \\
-    \hline
-    \end{tabular}
-    \end{table}
+\section{Conclusion}
+
+In this work, we explored the application of Structured State Space Sequence Models (SSMs), particularly the Mamba architecture, for tumor segmentation from brain MRI scans.
+The Mamba architecture leverages selective state spaces, enabling efficient and effective segmentation while maintaining linear scalability.
+
+Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
+The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) with a dice score of 0.8672.
+
+While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867.
+Notably, our Mamba-based UMambaBot\_PP and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
+
+We were able to make the ultra-light model smaller and more accurate on our task, further proving the potential of Mamba-based systems that prioritize compactness and efficiency.
+The selective state spaces and parallel processing strategies employed in our models enabled us to strike a balance between model size and segmentation accuracy.
+
+Overall, this work contributes to the exploration of Structured State Space Sequence Models for medical image segmentation tasks.
+The Mamba architecture's ability to capture long-range dependencies while maintaining linear scalability positions it as a promising alternative to traditional convolutional and transformer-based approaches, especially in resource-constrained environments.
+
+Future work could involve further refinements and optimizations of the Mamba-based models, as well as their application to other medical imaging modalities and segmentation tasks.
+Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.
 
@@
 \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
+\bibitem{ultralightvmunet} R.
+Wu, Y.
+Liu, P.
+Liang, and Q.
+Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr.
+20, 2024.
+[Online].
+Available: https://arxiv.org/pdf/2403.20035.pdf
+\bibitem{zhao2017pyramid}
+H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
 \end{thebibliography}

### 2024-04-21 paper/main.tex
add seg img to paper
update
update to have info on dataset

@@
 
+\section{Input Data}\label{sec2}
+We shall be using a Brain MRI segmentation dataset found here in \cite{dataset}.
+This dataset has been used in several papers regarding classification of the shape and severity of tumors, and provides an adequate dataset for research purposes in the field of medical image analysis, particularly in the area of brain MRI segmentation.
+Researchers have utilized this dataset in various studies focusing on the classification of tumor shapes and the assessment of tumor severity.
+With its comprehensive collection of brain MRI scans, along with corresponding segmentation masks, the dataset offers valuable resources for developing and evaluating algorithms aimed at automating tumor detection and analysis.
+
+The dataset contains brain MRI images along with their segmentation masks for 110 patients, for a total of almost 4000 total images, of which about 1400 have tumors.
+
+\section{Output Data}\label{sec3}
+The output of this model shall be segmentation masks for new, never before seen images.
+
 \section{Compared Models}
@@
 \end{figure}
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/masks.png}
+    \caption{Example segmentations.}
+    \label{fig:segs}
+\end{figure}
 
@@
 
-\section{Acknowledgment}
-
-This work was done as a final project for the University of South Carolina's CSCE763 Digital Image Processing course taught in Spring 2024 under Dr. Yan Tong.
-
 \begin{thebibliography}{00}
@@
 H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
+\bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
 \end{thebibliography}

### 2024-04-21 paper/main.tex
update paper

@@
 \usepackage{xcolor}
+\usepackage{hyperref}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{imgs/bars.png}
-    \caption{Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
-    \label{fig:bars}
-\end{figure}
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{imgs/balls.png}
-    \caption{A scatter plot of Dice Score vs model size in MegaBytes. The points are scaled to help represent size.}
-    \label{fig:balls}
-\end{figure}
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{imgs/masks.png}
-    \caption{Example segmentations.}
-    \label{fig:segs}
-\end{figure}
-
 \subsection{PP-UMamba}
@@
 The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters.
-According to the provided information, incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
+    \caption{Information about the train, test, and validate data.}
+    \label{fig:datastats}
+\end{figure}
 
 \section{Results}
-We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck, UNet, and UltraLight VM-UNet. The results are shown in Figures 1--2.
+We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck, UNet, and UltraLight VM-UNet.
+We train for 100 epochs on all models, and use dice-loss with Adam optimizer.
+We train on roughly 3000 images, and validate on roughly 400.
+we then test on roughly 400.
+We show the stats of our data in Figure \ref{fig:datastats}.
+The results are shown in Figures \ref{fig:bars},\ref{fig:balls},\ref{fig:segs}.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/bars.png}
+    \caption{Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
+    \label{fig:bars}
+\end{figure}
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/balls.png}
+    \caption{A scatter plot of Dice Score vs model size in MegaBytes. The points are scaled to help represent size.}
+    \label{fig:balls}
+\end{figure}
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/masks.png}
+    \caption{Example segmentations.}
+    \label{fig:segs}
+\end{figure}
 

### 2024-04-21 final_upload/tests/brain_mri_dataset.py

@@ -1,0 +1,54 @@
+##################################
+# Clay Crews, Lex Whalen
+#
+# Dataset class for Brain MRI
+# For use in later pipelines.
+
+# Necessary imports
+import torch
+from sklearn.model_selection import train_test_split
+import pandas as pd
+from glob import glob
+import cv2
+import numpy as np
+
+class BrainMRIDatasetBuilder:
+    def __init__(self, data_dir):
+        self.data_dir = data_dir
+
+    def create_df(self):
+        images_paths = []
+        masks_paths = glob(f'{self.data_dir}/*/*_mask*')
+
+        for i in masks_paths:
+            images_paths.append(i.replace('_mask', ''))
+
+        df = pd.DataFrame(data={'images_paths': images_paths, 'masks_paths': masks_paths})
+
+        return df
+
+    def split_df(self, df):
+        train_df, dummy_df = train_test_split(df, train_size=0.8, random_state=37)
+        valid_df, test_df = train_test_split(dummy_df, train_size=0.5, random_state=37)
+        return train_df, valid_df, test_df
+
+
+class BrainMRIDataset(torch.utils.data.Dataset):
+    def __init__(self, dataframe, transform=None, mask_transform=None):
+        self.df = dataframe
+        self.transform = transform
+        self.mask_transform = mask_transform
+
+    def __len__(self):
+        return len(self.df)
+
+    def __getitem__(self, idx):
+        image = cv2.imread(self.df.iloc[idx, 0]) / 255.0
+        mask = cv2.imread(self.df.iloc[idx, 1], cv2.IMREAD_GRAYSCALE) / 255.0
+        mask = np.where(mask >= 0.5, 1., 0.)
+        if self.transform:
+            image = self.transform(image)
+        if self.mask_transform:
+            mask = self.mask_transform(mask)
+
+        return image, mask

### 2024-04-21 final_upload/tests/calculate_dice_score.py

@@ -1,0 +1,16 @@
+import torch
+
+def calculate_dice_score(predicted_mask, target_mask):
+    # # Flatten predicted and target masks
+    # predicted_flat = predicted_mask.view(-1)
+    # target_flat = target_mask.view(-1)
+
+    # Calculate intersection
+    intersection = torch.sum(predicted_mask * target_mask)
+
+    # Calculate Dice coefficient
+    dice = (2. * intersection) / (torch.sum(predicted_mask) + torch.sum(target_mask))
+
+    
+
+    return dice

### 2024-04-21 final_upload/tests/calculate_iou.py

@@ -1,0 +1,9 @@
+import torch 
+
+def calculate_iou(pred_mask, true_mask):
+
+    intersection = torch.logical_and(pred_mask, true_mask).sum()
+    union = torch.logical_or(pred_mask, true_mask).sum()
+
+    iou = intersection / union if union != 0 else 0.0
+    return iou

### 2024-04-21 final_upload/tests/mamba_pipeline.py

@@ -1,0 +1,144 @@
+################################
+#
+# Clay Crews, Lex Whalen
+# Pyramidal U-Mamba
+# 
+# Explanation: 
+# We take the original U-Mamba architecture (see here: https://github.com/bowang-lab/U-Mamba) and add
+# Pyramidal Pooling 
+#
+import torch
+import torch.nn as nn
+from torch.utils.data import Dataset, DataLoader
+
+from mamba_model import UMambaBot
+from dice_loss import DiceLoss
+import torch.optim as optim
+from brain_mri_dataset import BrainMRIDatasetBuilder,BrainMRIDataset
+
+from transforms import BrainMRITransforms
+
+from calculate_iou import calculate_iou
+
+
+device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
+model = UMambaBot(
+    input_channels=3,  # Assuming RGB images with 3 channels
+    n_stages=5,
+    features_per_stage=(32, 64, 128, 256,512),
+    conv_op=nn.Conv2d,  # Assuming 2D convolution
+    kernel_sizes=(3, 3, 3, 3, 3),  # Adjusted kernel sizes for 2D convolution
+    strides=(1, 2, 2, 2, 2),
+    num_classes=3,
+    n_conv_per_stage=(1, 1, 1, 1, 1),
+    n_conv_per_stage_decoder=(1, 1, 1, 1),
+    conv_bias=True,
+    norm_op=nn.InstanceNorm2d,  # Assuming 2D instance normalization
+    norm_op_kwargs={},
+    dropout_op=None,
+    nonlin=nn.LeakyReLU,
+    nonlin_kwargs={'inplace': True},
+    # Pyramidal Pooling
+    ppm_pool_sizes=(1,2,3,6)
+).to(device)
+
+total_params = sum(p.numel() for p in model.parameters())
+trainable_params = sum(p.numel() for p in model.parameters() if p.requires_grad)
+print(f'Total parameters: {total_params}')
+print(f'Trainable parameters: {trainable_params}')
+
+learning_rate = 0.0003
+
+criterion = DiceLoss()
+optimizer = optim.Adam(
+    model.parameters(),
+    lr=learning_rate)
+
+epochs = 100
+
+train_loss = []
+val_loss = []
+trainIOU = []
+valIOU = []
+
+import sys
+
+builder = BrainMRIDatasetBuilder("../datasets/tumor_segs")
+df = builder.create_df()
[…]

### 2024-04-21 final_upload/tests/transforms.py

@@ -1,0 +1,17 @@
+##################################
+# Clay Crews, Lex Whalen
+#
+# Dataset class for Brain MRI Transforms.
+# For use in later pipelines.
+
+from torchvision import transforms
+
+class BrainMRITransforms:
+    def __init__(self, resize_shape=(256, 256)):
+        self.transforms = transforms.Compose([
+            transforms.ToTensor(),
+            transforms.Resize(resize_shape)
+        ])
+
+    def __call__(self, img):
+        return self.transforms(img)

### 2024-04-22 models/ResNet18.py
more models

@@ -1,0 +1,42 @@
+import torch
+import torch.nn as nn
+import torchvision.models as models
+
+
+
+def load_model_and_weights(device):
+    model = ResNet18Seg().to(device)
+
+    # Load weights for parallel training
+    state_dict = torch.load('resnet18_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+# ResNet18
+
+class ResNet18Seg(nn.Module):
+    def __init__(self, num_classes=1):
+        super(ResNet18Seg, self).__init__()
+        resnet = models.resnet18(pretrained=True)
+        self.encoder = nn.Sequential(*list(resnet.children())[:-2])  # Remove the final fully connected layer and avgpool layer
+        self.decoder = nn.Sequential(
+            nn.ConvTranspose2d(512, 256, kernel_size=4, stride=2, padding=1),  # Upsample to 8x8 spatial size
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(256, 128, kernel_size=4, stride=2, padding=1),  # Upsample to 16x16 spatial size
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(128, 64, kernel_size=4, stride=2, padding=1),   # Upsample to 32x32 spatial size
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(64, num_classes, kernel_size=4, stride=2, padding=1),  # Upsample to 64x64 spatial size (original input size)
+            nn.Upsample(size=(256, 256), mode='bilinear', align_corners=False)
+        )
+
+    def forward(self, x):
+        x = self.encoder(x)
+        x = self.decoder(x)
+        return x

### 2024-04-22 models/ResNet50.py

@@ -1,0 +1,41 @@
+import torch
+import torch.nn as nn
+import torchvision.models as models
+
+
+def load_model_and_weights(device):
+    model = ResNet50Seg().to(device)
+
+    # Load weights for parallel training
+    state_dict = torch.load('resnet50_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+# ResNet50
+
+class ResNet50Seg(nn.Module):
+    def __init__(self, num_classes=1):
+        super(ResNet50Seg, self).__init__()
+        resnet = models.resnet50(pretrained=True)
+        self.encoder = nn.Sequential(*list(resnet.children())[:-2])  # Remove the final fully connected layer and avgpool layer
+        self.decoder = nn.Sequential(
+            nn.ConvTranspose2d(2048, 512, kernel_size=4, stride=2, padding=1),
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(512, 256, kernel_size=4, stride=2, padding=1),
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(256, 128, kernel_size=4, stride=2, padding=1),
+            nn.ReLU(inplace=True),
+            nn.ConvTranspose2d(128, num_classes, kernel_size=4, stride=2, padding=1),
+            nn.Upsample(size=(256, 256), mode='bilinear', align_corners=False)
+        )
+
+    def forward(self, x):
+        x = self.encoder(x)
+        x = self.decoder(x)
+        return x

### 2024-04-22 models/UMambaBot.py

@@ -1,0 +1,285 @@
+import torch
+from torch import nn
+from mamba_ssm import Mamba
+
+import os
+from glob import glob
+import pandas as pd 
+import cv2
+import torch
+import torchvision
+from torchvision.models import vgg16_bn
+import torchvision.transforms as transforms
+from sklearn.model_selection import train_test_split
+import torch.nn as nn
+import numpy as np
+import matplotlib.pyplot as plt
+import numpy as np
+import torch
+from torch import nn
+from typing import Union, Type, List, Tuple
+
+from dynamic_network_architectures.building_blocks.helper import get_matching_convtransp
+from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
+from dynamic_network_architectures.building_blocks.plain_conv_encoder import PlainConvEncoder
+from dynamic_network_architectures.building_blocks.residual import StackedResidualBlocks
+
+from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
+from dynamic_network_architectures.building_blocks.residual import BasicBlockD, BottleneckD
+from torch.nn.modules.conv import _ConvNd
+from torch.nn.modules.dropout import _DropoutNd
+from dynamic_network_architectures.building_blocks.helper import convert_conv_op_to_dim
+
+from nnunetv2.utilities.plans_handling.plans_handler import ConfigurationManager, PlansManager
+from dynamic_network_architectures.building_blocks.helper import get_matching_instancenorm, convert_dim_to_conv_op
+from nnunetv2.utilities.network_initialization import InitWeights_He
+from mamba_ssm import Mamba
+
+
+## Custom class imports
+#
+
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+# Ignore Warnings
+import warnings
+warnings.filterwarnings("ignore")
+
+
+
+
+
+def load_model_and_weights(device):
+    model = UMambaBot(
+        input_channels=3,  # Assuming RGB images with 3 channels
+        n_stages=5,
+        features_per_stage=(32, 64, 128, 256,512),
+        conv_op=nn.Conv2d,  # Assuming 2D convolution
+        kernel_sizes=(3, 3, 3, 3, 3),  # Adjusted kernel sizes for 2D convolution
+        strides=(1, 2, 2, 2, 2),
+        num_classes=1,
+        n_conv_per_stage=(1, 1, 1, 1, 1),
[…]

### 2024-04-22 models/UMambaBot_pp.py

@@ -1,0 +1,288 @@
+import torch
+from torch import nn
+from mamba_ssm import Mamba
+
+import os
+from glob import glob
+import pandas as pd 
+import cv2
+import torch
+import torchvision
+from torchvision.models import vgg16_bn
+import torchvision.transforms as transforms
+from sklearn.model_selection import train_test_split
+import torch.nn as nn
+import numpy as np
+import matplotlib.pyplot as plt
+import numpy as np
+import torch
+from torch import nn
+from typing import Union, Type, List, Tuple
+
+from dynamic_network_architectures.building_blocks.helper import get_matching_convtransp
+from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
+from dynamic_network_architectures.building_blocks.plain_conv_encoder import PlainConvEncoder
+from dynamic_network_architectures.building_blocks.residual import StackedResidualBlocks
+
+from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
+from dynamic_network_architectures.building_blocks.residual import BasicBlockD, BottleneckD
+from torch.nn.modules.conv import _ConvNd
+from torch.nn.modules.dropout import _DropoutNd
+from dynamic_network_architectures.building_blocks.helper import convert_conv_op_to_dim
+
+from nnunetv2.utilities.plans_handling.plans_handler import ConfigurationManager, PlansManager
+from dynamic_network_architectures.building_blocks.helper import get_matching_instancenorm, convert_dim_to_conv_op
+from nnunetv2.utilities.network_initialization import InitWeights_He
+from mamba_ssm import Mamba
+
+
+## Custom class imports
+#
+
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+# Ignore Warnings
+import warnings
+warnings.filterwarnings("ignore")
+
+
+
+
+
+def load_model_and_weights(device):
+    model = UMambaBot(
+        input_channels=3,  # Assuming RGB images with 3 channels
+        n_stages=5,
+        features_per_stage=(32, 64, 128, 256,512),
+        conv_op=nn.Conv2d,  # Assuming 2D convolution
+        kernel_sizes=(3, 3, 3, 3, 3),  # Adjusted kernel sizes for 2D convolution
+        strides=(1, 2, 2, 2, 2),
+        num_classes=1,
+        n_conv_per_stage=(1, 1, 1, 1, 1),
[…]

### 2024-04-22 models/UNet.py

@@ -1,0 +1,142 @@
+import torch
+import torch.nn as nn
+
+
+def load_model_and_weights(device):
+    model = UNet().to(device)
+
+    # Load weights 
+    state_dict = torch.load('unet_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+
+class UNet(nn.Module):
+    def __init__(self):
+        super(UNet, self).__init__()
+
+        # Encoder
+        self.encoder1 = nn.Sequential(
+            nn.Conv2d(3, 32, kernel_size=3, padding=1),
+            nn.BatchNorm2d(32),
+            nn.ReLU(inplace=True),
+            nn.Conv2d(32, 32, kernel_size=3, padding=1),
+            nn.BatchNorm2d(32),
+            nn.ReLU(inplace=True)
+        )
+        self.pool1 = nn.MaxPool2d(kernel_size=2, stride=2)
+        self.encoder2 = nn.Sequential(
+            nn.Conv2d(32, 64, kernel_size=3, padding=1),
+            nn.BatchNorm2d(64),
+            nn.ReLU(inplace=True),
+            nn.Conv2d(64, 64, kernel_size=3, padding=1),
+            nn.BatchNorm2d(64),
+            nn.ReLU(inplace=True)
+        )
+        self.pool2 = nn.MaxPool2d(kernel_size=2, stride=2)
+        self.encoder3 = nn.Sequential(
+            nn.Conv2d(64, 128, kernel_size=3, padding=1),
+            nn.BatchNorm2d(128),
+            nn.ReLU(inplace=True),
+            nn.Conv2d(128, 128, kernel_size=3, padding=1),
+            nn.BatchNorm2d(128),
+            nn.ReLU(inplace=True)
+        )
+        self.pool3 = nn.MaxPool2d(kernel_size=2, stride=2)
+        self.encoder4 = nn.Sequential(
+            nn.Conv2d(128, 256, kernel_size=3, padding=1),
+            nn.BatchNorm2d(256),
+            nn.ReLU(inplace=True),
+            nn.Conv2d(256, 256, kernel_size=3, padding=1),
+            nn.BatchNorm2d(256),
+            nn.ReLU(inplace=True)
+        )
[…]

### 2024-04-22 models/segformer.py

@@ -1,0 +1,518 @@
+import numpy as np
+import torch
+import torchvision
+
+import warnings
+warnings.filterwarnings('ignore')
+
+import os
+import gc
+from typing import List, Iterable
+
+from IPython.display import clear_output
+gc.enable()
+
+import torch
+import torch.nn as nn
+from torch.utils.data import Dataset, DataLoader
+from torchvision.ops import StochasticDepth
+
+from einops import rearrange
+
+import numpy as np
+import pandas as pd
+
+import matplotlib.pyplot as plt
+import cv2
+from PIL import Image, ImageDraw
+
+from sklearn.model_selection import train_test_split
+
+from torchvision import transforms
+
+import albumentations as A
+from albumentations.pytorch.transforms import ToTensorV2
+
+from brain_mri_dataset import BrainMRIDatasetBuilder,BrainMRIDataset
+
+from glob import glob
+
+seed = 37
+pd.set_option('display.max_colwidth', None)
+
+device = 'cuda' if torch.cuda.is_available() else 'cpu'
+
+def load_data_model_and_weights(train_df, val_df, test_df):
+
+    train_data = SegDataset(train_df, transforms= train_transforms)
+    val_data = SegDataset(val_df, transforms= valid_transforms)
+    test_data = SegDataset(test_df, transforms= valid_transforms)
+
+    train_dataloader = DataLoader(train_data, batch_size = BATCH_SIZE , shuffle = True)
+    val_dataloader = DataLoader(val_data, batch_size = BATCH_SIZE , shuffle = False)
+    test_dataloader = DataLoader(test_data, batch_size = BATCH_SIZE , shuffle = False)
+
+    model = nn.DataParallel(SegFormer(
+                    in_channels=3,
+                    widths=[64, 128, 256, 512],
+                    depths=[3, 4, 6, 3],
+                    all_num_heads=[1, 2, 4, 8],
+                    patch_sizes=[7, 3, 3, 3],
+                    overlap_sizes=[4, 2, 2, 2],
+                    reduction_ratios=[8, 4, 2, 1],
+                    mlp_expansions=[4, 4, 4, 4],
+                    decoder_channels=256,
+                    scale_factors=[8, 4, 2, 1],
+                    num_classes=NUM_CLASSES,
+                                            )).to(device)
+
+    
+    # Load weights for parallel training
+    state_dict = torch.load('segformer_v2_weights.pth', map_location=device)
[…]

### 2024-04-22 models/ultralightVMUNet.py

@@ -1,0 +1,269 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+
+
+def load_model_and_weights(device):
+    model = nn.DataParallel(UltraLight_VM_UNet(
+            num_classes=1,
+            input_channels=3,
+            c_list=[8,16,24,32,48,64],
+            split_att='fc',
+            bridge=True
+        )).to(device)
+    
+    # Load weights for parallel training
+    state_dict = torch.load('ultralightmunet_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    # state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//4, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+        B, C = x.shape[:2]
+        assert C == self.input_dim
+        n_tokens = x.shape[2:].numel()
+        img_dims = x.shape[2:]
+        x_flat = x.reshape(B, C, n_tokens).transpose(-1, -2)
+        x_norm = self.norm(x_flat)
+
+        x1, x2, x3, x4 = torch.chunk(x_norm, 4, dim=2)
+        x_mamba1 = self.mamba(x1) + self.skip_scale * x1
[…]

### 2024-04-22 models/ultralightVMUNet_v1.py

@@ -1,0 +1,256 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+
+
+def load_model_and_weights(device):
+    # 32, 64, 128, 256, 512
+    model = UltraLight_VM_UNet_v1(
+        num_classes=1,
+        input_channels=3,
+        # c_list=[8,16,24,32,48,64],
+        c_list=[16, 32, 64, 128, 256],
+        split_att='fc',
+        bridge=True
+    ).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//4, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+        B, C = x.shape[:2]
+        assert C == self.input_dim
+        n_tokens = x.shape[2:].numel()
+        img_dims = x.shape[2:]
+        x_flat = x.reshape(B, C, n_tokens).transpose(-1, -2)
+        x_norm = self.norm(x_flat)
+
+        x1, x2, x3, x4 = torch.chunk(x_norm, 4, dim=2)
[…]

### 2024-04-22 models/ultralightVMUNet_v2.py

@@ -1,0 +1,270 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+
+def load_model_and_weights(device):
+    model = UltraLight_VM_UNet_v2(
+        num_classes=1,
+        input_channels=3,
+        c_list=[8,16,24,32,48,64],
+        split_att='fc',
+        bridge=True
+    ).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_v2_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//8, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+        B, C = x.shape[:2]
+        assert C == self.input_dim
+        n_tokens = x.shape[2:].numel()
+        img_dims = x.shape[2:]
+        x_flat = x.reshape(B, C, n_tokens).transpose(-1, -2)
+        x_norm = self.norm(x_flat)
+
+        x1, x2, x3, x4, x5, x6, x7, x8 = torch.chunk(x_norm, 8, dim=2)
+        x_mamba1 = self.mamba(x1) + self.skip_scale * x1
[…]

### 2024-04-22 models/ultralightVMUNet_v3.py

@@ -1,0 +1,258 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+
+def load_model_and_weights(device):
+    # 32, 64, 128, 256, 512
+    model = UltraLight_VM_UNet_v3(
+        num_classes=1,
+        input_channels=3,
+        # c_list=[8,16,24,32,48,64],
+        c_list=[16, 32, 64, 128, 256],
+        split_att='fc',
+        bridge=True
+    ).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_v3_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//8, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+        B, C = x.shape[:2]
+        assert C == self.input_dim
+        n_tokens = x.shape[2:].numel()
+        img_dims = x.shape[2:]
+        x_flat = x.reshape(B, C, n_tokens).transpose(-1, -2)
+        x_norm = self.norm(x_flat)
+
+        x1, x2, x3, x4, x5, x6, x7, x8 = torch.chunk(x_norm, 8, dim=2)
[…]

### 2024-04-22 models/ultralightVMUNet_v4.py

@@ -1,0 +1,280 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+
+def load_model_and_weights(device):
+    model = nn.DataParallel(UltraLight_VM_UNet(
+        num_classes=1,
+        input_channels=3,
+        # c_list=[8,16,24,32,48,64],
+        c_list=[16,32,64,128,256,512],
+        split_att='fc',
+        bridge=True
+    )).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_v4_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    # state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//8, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+        self.ppm = PyramidalPoolingModule(input_channels=input_dim,
+                                          pool_sizes=(1,2,3,6),
+                                          output_channels=[input_dim//8, input_dim//4, input_dim//2, input_dim//8]) 
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+
[…]

### 2024-04-22 models/ultralightVMUNet_v5.py

@@ -1,0 +1,287 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+
+def load_model_and_weights(device):
+    model = nn.DataParallel(UltraLight_VM_UNet(
+        num_classes=1,
+        input_channels=3,
+        # c_list=[8,16,24,32,48,64],
+        c_list=[16,32,64,128,256,512],
+        split_att='fc',
+        bridge=True
+    )).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_v5_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    # state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//8, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+    
+    def forward(self, x):
+        if x.dtype == torch.float16:
+            x = x.type(torch.float32)
+
+        B, C = x.shape[:2]
[…]

### 2024-04-22 tests/calculate_dice_score.py

@@ -11,6 +11,8 @@ def calculate_dice_score(predicted_mask, target_mask):
     # Calculate Dice coefficient
     dice = (2. * intersection) / (torch.sum(predicted_mask) + torch.sum(target_mask))
 
-    
+    # Check if dice is a tensor
+    if torch.is_tensor(dice):
+        dice = dice.item()  # Convert tensor to Python float
 
     return dice

### 2024-04-22 tests/calculate_iou.py

@@ -6,4 +6,9 @@ def calculate_iou(pred_mask, true_mask):
     union = torch.logical_or(pred_mask, true_mask).sum()
 
     iou = intersection / union if union != 0 else 0.0
+
+    # Check if iou is a tensor
+    if torch.is_tensor(iou):
+        iou = iou.item()  # Convert tensor to Python float
+
     return iou

### 2024-04-22 paper/main.tex
citation
small update

@@
 
+\section{Proposed Methodology}
+In this paper, we strive to maintain high accuracy in our image segmentation while also keeping low parameter counts.
+We primarily seek to use pyramidal pooling modules \cite{zhao2017pyramid} to improve accuracy, and use a variant of the Mamba \cite{gu2023mamba} architecture inspired by \cite{ultralightvmunet} to maintain small parameter sizes.
+
 \section{Input Data}\label{sec2}
@@
 
-The dataset contains brain MRI images along with their segmentation masks for 110 patients, for a total of almost 4000 total images, of which about 1400 have tumors.
+The dataset contains brain MRI images along with their segmentation masks for 110 patients.
+There were a total of 3143 train images, 393 validation images, and 393 test images.
+The distribution of the tumor and non-tumor images is shown in Figure \ref{fig:datastats}.
 
@@
 The output of this model shall be segmentation masks for new, never before seen images. 
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
+    \caption{Information about the train, test, and validate data.}
+    \label{fig:datastats}
+\end{figure}
+
+\section{Mamba}
+Before we explain the compared models, we first give a brief overview of Mamba. [to do]
 
@@
 For this our analysis, we compare four models against our's.
-The models are U-Net\cite{ronneberger2015unet},ResNet18, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
+The models are U-Net\cite{ronneberger2015unet},ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
 Below we explain our choices.
@@
 Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
-    \caption{Information about the train, test, and validate data.}
-    \label{fig:datastats}
-\end{figure}
 

### 2024-04-23 paper/main.tex
update paper

@@
 U-net and transformer-based architectures have dominated medical image segmentation, particularly for tumor segmentation from brain MRI scans.
-However, Structured State Space Sequence Models (SSMs), as highlighted by Gu, Goel, and Ré, present a promising alternative with comparable modeling power to transformers but linear scalability in sequence length, making them well-suited for large medical images.
+However, recent advances in Structured State Space Sequence Models (SSMs) present a promising alternative with comparable modeling power to transformers but linear scalability in sequence length, making them well-suited for large medical images.
 
@@
 We will evaluate the model's performance using standard segmentation metrics and compare the results with U-net and transformer-based approaches.
+We open-source the code here: \href{https://github.com/lxaw/mamba-tumor-seg}{https://github.com/lxaw/mamba-tumor-seg}.
 \end{abstract}
@@
 \section{Introduction}
-MRI imaging to identify brain tumors has naturally been a target for accurate image segmentation through the use of ML architecture.
+MRI imaging to identify brain tumors has naturally been a target for accurate image segmentation through the use of machine learning (ML) architecture.
 The use of computer-aided diagnosis of tumors can be crucial to quickly identify tumors and determine a course of action for the patient.
@@
 \section{Input Data}\label{sec2}
-We shall be using a Brain MRI segmentation dataset found here in \cite{dataset}.
+We shall be using a Brain MRI segmentation dataset found in \cite{dataset}.
 This dataset has been used in several papers regarding classification of the shape and severity of tumors, and provides an adequate dataset for research purposes in the field of medical image analysis, particularly in the area of brain MRI segmentation.
@@
 \section{Mamba}
-Before we explain the compared models, we first give a brief overview of Mamba. [to do]
+Mamba \cite{gu2023mamba} is a recently proposed architecture that combines the modeling power of Transformers with the linear scaling efficiency of state space models (SSMs).
+The key idea is to augment SSMs with a selection mechanism that allows the model parameters to vary based on the input, enabling it to perform content-aware reasoning.
+Standard SSMs map an input sequence $x(t)$ to an output $y(t)$ via a latent state $h(t)$ using a linear time-invariant system:
+\begin{align*}
+h'(t) &= \mathbf{A}h(t) + \mathbf{B}x(t) \\
+y(t) &= \mathbf{C}h(t) + \mathbf{D}x(t)
+\end{align*}
+where the parameter matrices $\mathbf{A}$, $\mathbf{B}$, $\mathbf{C}$, $\mathbf{D}$ are fixed.
+In the original paper, the $\mathbf{D}$ matrix is considered only as a skip connection, and thus as it does not directly play a role in the differential equation, was ignored.
+This allows SSMs to be computed efficiently as a convolution.
+However, the time-invariance means the model cannot change its behavior for specific inputs, making it difficult to solve tasks requiring selective focus, such as selective copying or induction heads \cite{gu2023mamba}.
+Mamba introduces a selection mechanism where the SSM parameters $\mathbf{B}$, $\mathbf{C}$, $\Delta$ (the state step size) vary for each input token $x_t$.
+This allows the model to selectively propagate or forget information based on the current token.
+The selective SSM is computed using a parallel scan operation to maintain linear complexity.
+The Mamba block interleaves the selective SSM with linear projections, convolutions, and activations:
+\begin{enumerate}
+\item Linear projection to expand the embedding dimension
+\item Convolution to mix information between dimensions
+\item Selective SSM via parallel scan to efficiently propagate information
+\item Linear projection to get back to the target embedding size
+\end{enumerate}
+These blocks are stacked to form the Mamba architecture.
+Compared to Transformers, Mamba can achieve comparable modeling performance while scaling linearly in sequence length rather than quadratically \cite{gu2023mamba}.
+Overall, the selective SSM leverages the strengths of RNNs, CNNs and Transformers - it maintains an unbounded context window like RNNs, can be parallelized like CNNs, and achieves content-aware reasoning like Transformers, while being more efficient than all three.
+This makes Mamba a promising foundation model for processing long sequences across various domains.
 
@@
 
-For this our analysis, we compare four models against our's.
+For this our analysis, we compare four models against our own developed ones.
 The models are U-Net\cite{ronneberger2015unet},ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
@@
 These upsampling layers aim to increase the resolution of the output.
-Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information [1].
+Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
  A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part.
@@
 As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture.
-Unlike traditional convolutional networks, U-Net does not employ any fully connected layers and only utilizes the valid part of each convolution [2].
 To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image.
@@
 U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures.
-The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) (Gu et al., 2022), a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies.
+The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies.
 By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks.
@@
 In the ATM module, the similarity maps between a set of learnable class tokens and the spatial feature maps are transferred to the segmentation masks, enabling effective semantic segmentation.
-SegViT has demonstrated impressive performance on several benchmark datasets, including ADE20K, COCO-Stuff-10K, and PASCAL-Context, achieving state-of-the-art results.
 Moreover, to address the computational cost associated with the ViT backbone, SegViT proposes query-based down-sampling (QD) and query-based up-sampling (QU) to build a Shrunk structure.
@@
     \includegraphics[width=\columnwidth]{imgs/balls.png}
-    \caption{A scatter plot of Dice Score vs model size in MegaBytes. The points are scaled to help represent size.}
+    \caption{A scatter plot of Dice Score vs model parameter count, where "M" means "millions. The points are scaled to help represent size.}
     \label{fig:balls}
@@
 Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
-The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) with a dice score of 0.8672.
+The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
 

### 2024-04-23 paper/main.tex
comments and suggestions

@@
 \usepackage{hyperref}
+
+\newcommand{\gr}[1]{\textcolor{green}{#1}}
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
     \includegraphics[width=\columnwidth]{imgs/data_stats.png}
-    \caption{Information about the train, test, and validate data.}
+    \caption{Distributions of tumor and non-tumor images in the train, validation, and test datasets.}
     \label{fig:datastats}
@@
 \begin{align*}
-h'(t) &= \mathbf{A}h(t) + \mathbf{B}x(t) \\
-y(t) &= \mathbf{C}h(t) + \mathbf{D}x(t)
+    h'(t) & = \mathbf{A}h(t) + \mathbf{B}x(t) \\
+    y(t)  & = \mathbf{C}h(t) + \mathbf{D}x(t)
 \end{align*}
+\gr{Do these need to be numbered equations? -- I don't. Nvm, disregard unless you think so}
+
 where the parameter matrices $\mathbf{A}$, $\mathbf{B}$, $\mathbf{C}$, $\mathbf{D}$ are fixed.
@@
 For this our analysis, we compare four models against our own developed ones.
-The models are U-Net\cite{ronneberger2015unet},ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
+The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
 Below we explain our choices.
@@
 \subsection{SegViT}
+\gr{We used SegFormer?}
 SegViT \cite{zhang2022segvit} explores the capability of plain Vision Transformers (ViTs) \cite{dosovitskiy2021image} for semantic segmentation, leveraging the attention mechanism to generate masks for semantic segmentation.
@@
 \subsection{PP-UMamba}
+
+\gr{To address the suggestion from Tong for "an overview of the whole pipeline" we should probably explain how PP is incorporated.
+Either by describing the pipeline or making a graphic for the pipeline.}
+
 This model incorporates pyramidal pooling, which is a strategy used in convolutional neural networks (CNNs) to capture context and incorporate multi-scale information \cite{zhao2017pyramid}.
@@
 In our PP-UMamba model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
+
+% Unsure why this comment line is giving me troubles !!
+\gr{$These should all be UL-VM-UNet, or ULVM_UNet if you are against the extra dash. UltraLight VisionMamba-UNet This in each subsection title and in the paragraphs.$}
 
@@
 \subsection{ULMUNet\_v4}
+\gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks.
+If we want to do similar to PP-UMamba we can do it just in the bottleneck.
+Let me know if we want this can I can do it fast.}
 This model incorporates pyramidal pooling, similar to PP-UMamba, but within the UltraLight VM-UNet architecture.
@@
 \subsection{UL-PP}
+\gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
 The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters.
@@
 \section{Results}
-We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck, UNet, and UltraLight VM-UNet.
+We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck \gr{UMamba bottleneck rather than "with"}, UNet, and UltraLight VM-UNet.
 We train for 100 epochs on all models, and use dice-loss with Adam optimizer.
@@
     \includegraphics[width=\columnwidth]{imgs/bars.png}
-    \caption{Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
+    \caption{\gr{Thoughts on this vs the two bar graph?}Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
     \label{fig:bars}
 \end{figure}
+
 \begin{figure}[!t]
@@
 \end{figure}
+\gr{do we need a table of the params and scores? May be helpful.. Maybe stick one in the appendix. }
+
 \begin{figure}[!t]
@@
 \end{figure}
+\gr{let's include more segmentation results. Maybe in appendix.}
+
+\gr{Would appendix really need to be in the page limit? Appendix is to include figures that wouldn't fit in paper?}
 
@@
 Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
-The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
+The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) \gr{how does this outperform?
+Dice score is lower.}with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
 
-While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867.
-Notably, our Mamba-based UMambaBot\_PP and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
+While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}.
+Notably, our Mamba-based PP-UMambaBot \gr{name the section in OUR MODELS PP-UMambaBot or change this one.
+Consistency.} and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
 
@@
     \bibitem{zhao2017pyramid}
-H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
+    H.
+Zhao, J.
+Shi, X.
+Qi, X.
+Wang and J.
+Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp.
+6230-6239, doi: 10.1109/CVPR.2017.660.
     \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation

### 2024-04-24 paper/main.tex
start

@@
 \end{align*}
-\gr{Do these need to be numbered equations? -- I don't. Nvm, disregard unless you think so}
 
@@
 Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
-\subsection{SegViT}
-\gr{We used SegFormer?}
-SegViT \cite{zhang2022segvit} explores the capability of plain Vision Transformers (ViTs) \cite{dosovitskiy2021image} for semantic segmentation, leveraging the attention mechanism to generate masks for semantic segmentation.
-Unlike previous ViT-based segmentation networks that learn pixel-level representations from the output of the ViT, SegViT introduces the Attention-to-Mask (ATM) module.
-In the ATM module, the similarity maps between a set of learnable class tokens and the spatial feature maps are transferred to the segmentation masks, enabling effective semantic segmentation.
-Moreover, to address the computational cost associated with the ViT backbone, SegViT proposes query-based down-sampling (QD) and query-based up-sampling (QU) to build a Shrunk structure.
-By comparing U-Mamba against SegViT, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the attention-based segmentation mechanism employed by SegViT.
+\subsection{SegFormer}
+SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA).
+The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks.
+The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution.
+This allows SegFormer to efficiently process high-resolution images while capturing both local and global context.
+Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details.
+By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer.
 This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks.
-Furthermore, it will help us understand the potential of attention mechanisms and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.
+Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.
 
@@
 
-\subsection{PP-UMamba}
-
-\gr{To address the suggestion from Tong for "an overview of the whole pipeline" we should probably explain how PP is incorporated.
-Either by describing the pipeline or making a graphic for the pipeline.}
+\subsection{UMambaBot\_PPBot}
 
@@
 
-In our PP-UMamba model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
+In our UMambaBot\_PP model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
 
 % Unsure why this comment line is giving me troubles !!
-\gr{$These should all be UL-VM-UNet, or ULVM_UNet if you are against the extra dash. UltraLight VisionMamba-UNet This in each subsection title and in the paragraphs.$}
 
-\subsection{ULMUNet\_v1}
+\subsection{UL-VM-UNet\_v1}
 For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder.
@@
 
-\subsection{ULMUNet\_v2}
+\subsection{UL-VM-UNet\_v2}
 In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8.
@@
 
-\subsection{ULMUNet\_v3}
+\subsection{UL-VM-UNet\_v3}
 ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2.
@@
 
-\subsection{ULMUNet\_v4}
+\subsection{UL-VM-UNet\_v4}
 \gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks.
-If we want to do similar to PP-UMamba we can do it just in the bottleneck.
+If we want to do similar to UMambaBot\_PP we can do it just in the bottleneck.
 Let me know if we want this can I can do it fast.}
-This model incorporates pyramidal pooling, similar to PP-UMamba, but within the UltraLight VM-UNet architecture.
+This model incorporates pyramidal pooling, similar to UMambaBot\_PP, but within the UltraLight VM-UNet architecture.
 The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.
 
-\subsection{UL-PP}
+\subsection{UL-VM-UNet\_v5}
 \gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
@@
 \section{Results}
-We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck \gr{UMamba bottleneck rather than "with"}, UNet, and UltraLight VM-UNet.
+We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet.
 We train for 100 epochs on all models, and use dice-loss with Adam optimizer.
@@
 
-\begin{figure}[!t]
+\begin{table}[ht]
     \centering
-    \includegraphics[width=\columnwidth]{imgs/bars.png}
-    \caption{\gr{Thoughts on this vs the two bar graph?}Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
-    \label{fig:bars}
-\end{figure}
+    \begin{tabular}{|l|c|c|}
+        \hline
+        \textbf{Model Name} & \textbf{Parameter Count (Millions)} & \textbf{Dice Score} \\
+        \hline
+        UMambaBot\_PP & 10.0 & 0.8867 \\
+        UMambaBot & 9.8 & 0.8799 \\
+        ResNet18 & 13.9 & 0.8648 \\
+        ResNet50 & 42.9 & 0.8672 \\
+        SegFormer & 17.8 & 0.8454 \\
+        UNet & 7.8 & 0.8929 \\
+        UL-VM-UNet & 0.049 & 0.79955 \\
+        UL-VM-UNet\_v1 & 0.20 & 0.8409 \\
+        UL-VM-UNet\_v2 & 0.042 & 0.8243 \\
+        UL-VM-UNet\_v3 & 0.176 & 0.8109 \\
+        UL-VM-UNet\_v4 & 1.1 & 0.8548 \\
+        UL-VM-UNet\_v5 & 0.7 & 0.8633 \\
+        \hline
+    \end{tabular}
+    \caption{Comparison of different segmentation models.}
+    \label{tab:model_comparison}
+\end{table}
 
@@
 \end{figure}
-\gr{let's include more segmentation results. Maybe in appendix.}
 
@@
 Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
-The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) \gr{how does this outperform?
-Dice score is lower.}with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
+The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
 
-While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}.
-Notably, our Mamba-based PP-UMambaBot \gr{name the section in OUR MODELS PP-UMambaBot or change this one.
-Consistency.} and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
+While the traditional U-Net model achieved the highest dice score of 0.8929, our UMambaBot\_PP model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}.
+Notably, our Mamba-based UMambaBot\_PPBot and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
 
@@
 Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.
+\appendix
+% Redefine the section numbering for the appendix (optional)
+\renewcommand{\thesection}{\Alph{section}}
+% First appendix section
+\subsection{Training Pipeline}
+The training pipeline for our model consists of the following steps:
+
+1. Data preprocessing: The dataset is split into training, validation, and test sets. The images are resized to a uniform size and normalized.
+
+2. Model architecture: We employ a convolutional neural network (CNN) with the following layers: [Describe the layers and their configurations].
+
+3. Training: The model is trained using the training set for [X] epochs with a batch size of [Y]. We use the [Z] optimizer with a learning rate of [L].
+
+4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use metrics such as accuracy, precision, recall, and F1-score.
+
+5. Testing: Once the model achieves satisfactory performance on the validation set, it is tested on the held-out test set to assess its generalization ability.
+
+% Second appendix section
+\subsection{Supplementary Figures}
 
@@
     \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
+    \bibitem{xie2021segformer} E.
+Xie, W.
+Wang, Z.
+Yu, A.
+Anandkumar, J.
+Alvarez, and P.
+Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
 \end{thebibliography}

### 2024-04-25 paper/main.tex
paper edits

@@
 \usepackage{hyperref}
+\usepackage{array}
+\usepackage{caption} % for \captionsetup command
+\usepackage{svg}
 
@@
 In this paper, we strive to maintain high accuracy in our image segmentation while also keeping low parameter counts.
-We primarily seek to use pyramidal pooling modules \cite{zhao2017pyramid} to improve accuracy, and use a variant of the Mamba \cite{gu2023mamba} architecture inspired by \cite{ultralightvmunet} to maintain small parameter sizes.
+We primarily seek to use pyramidal pooling modules \cite{zhao2017pyramid} to improve accuracy, and use a variant of the Mamba \cite{gu2023mamba} architecture inspired by \cite{wu2024ultralight} to maintain small parameter sizes.
 
@@
 For this our analysis, we compare four models against our own developed ones.
-The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}.
+The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{he2015deep}, ResNet50, U-Mamba\cite{ma2024umamba}, and SegViT\cite{dosovitskiy2021image}.
 Below we explain our choices.
@@
 Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models.
-For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{DBLP:journals/corr/abs-2006-11239} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}.
+For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{ho2020denoising} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}.
 We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.
@@
 These architectures, introduced by He et al.
-in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks.
+in their seminal work "Deep Residual Learning for Image Recognition" \cite{he2015deep}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks.
 The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks.
@@
 \subsection{U-Mamba}
-As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison.
+As our model takes much inspiration from the recently developed U-Mamba design \cite{ma2024umamba}, we also choose to incorporate it in our comparison.
 U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures.
@@
 \subsection{UltraLight VM-UNet}
-The UltraLight VM-UNet \cite{ultralightvmunet} is a lightweight neural network architecture designed for skin lesion segmentation tasks.
+The UltraLight VM-UNet \cite{wu2024ultralight} is a lightweight neural network architecture designed for skin lesion segmentation tasks.
 It is built upon the Vision Mamba module, which is a state-space model (SSM) that can efficiently handle long-range dependencies in sequences, making it well-suited for image segmentation tasks.
@@
 
-Inspired by \cite{U-Mamba} and \cite{ultralightvmunet}, we sought to implement both models on a different problem: tumor segmentation.
+Inspired by \cite{ma2024umamba} and \cite{wu2024ultralight}, we sought to implement both models on a different problem: tumor segmentation.
 
@@
 
-\subsection{UMambaBot\_PPBot}
+\subsection{UMambaBot-PP}
 
@@
 
-In our UMambaBot\_PP model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
+In our UMambaBot-PP model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.
 
@@
 
-\subsection{UL-VM-UNet\_v1}
-For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder.
+\subsection{UL-VM-UNet-v1}
+For UL-VM-UNet-v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder.
 Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step.
@@
 
-\subsection{UL-VM-UNet\_v2}
-In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8.
-This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.
+\subsection{UL-VM-UNet-v2}
+In UL-VM-UNet-v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8.
+This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{wu2024ultralight} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.
 
-\subsection{UL-VM-UNet\_v3}
-ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2.
+\subsection{UL-VM-UNet-v3}
+UL-VM-UNet-v3 combines the modifications from UL-VM-UNet-v1 and UL-VM-UNet-v2.
 It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256].
@@
 
-\subsection{UL-VM-UNet\_v4}
-\gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks.
-If we want to do similar to UMambaBot\_PP we can do it just in the bottleneck.
-Let me know if we want this can I can do it fast.}
-This model incorporates pyramidal pooling, similar to UMambaBot\_PP, but within the UltraLight VM-UNet architecture.
+\subsection{UL-VM-UNet-v4}
+This model incorporates pyramidal pooling, similar to UMambaBot-PP, but within every PVM block in both the encoder and decoder in the UltraLight VM-UNet architecture.
 The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.
 
-\subsection{UL-VM-UNet\_v5}
-\gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
-The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters.
+\subsection{UL-VM-UNet-v5}
+UL-VM-UNet-v5 is meant to be a comparison to the -v4 and -v6 UltraLight variations.
+This model trained and performed well with 1.1M parameters.
 Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+
+\subsection{UL-VM-UNet-v6}
+Another variation to further test the effects of incorporating pyramidal pooling in the UltraLight VM-UNet architecture was made by only using pyramidal pooling in the bottleneck layer, similarly done in UMambaBot-PP.
+The main goal here, again, was to integrate pyramidal pooling for the benefit of multi-scale feature extraction while keeping a low parameter count.
 
@@
 we then test on roughly 400.
-We show the stats of our data in Figure \ref{fig:datastats}.
-The results are shown in Figures \ref{fig:bars},\ref{fig:balls},\ref{fig:segs}.
+We show the distributions of our data in Figure \ref{fig:datastats}.
+Model results are shown in Table \ref{tab:model_comparison} and Figures \ref{fig:balls},\ref{fig:segs}, with additional segmentation results in Appendix \ref{supp_figs}.
 
+% Adjust the vertical spacing between rows
+\renewcommand{\arraystretch}{1.25}
 \begin{table}[ht]
     \centering
+    \captionsetup{skip=6pt}
     \begin{tabular}{|l|c|c|}
         \hline
-        \textbf{Model Name} & \textbf{Parameter Count (Millions)} & \textbf{Dice Score} \\
+        \textbf{Model Name} & \textbf{\#Params (M) $\downarrow$} & \textbf{Dice Score $\uparrow$} \\
         \hline
-        UMambaBot\_PP & 10.0 & 0.8867 \\
-        UMambaBot & 9.8 & 0.8799 \\
+        UMambaBot-PP(Ours) & 10.0 & 0.8867 \\
+        UMambaBot & \textbf{9.8} & 0.8799 \\
         ResNet18 & 13.9 & 0.8648 \\
@@
         SegFormer & 17.8 & 0.8454 \\
-        UNet & 7.8 & 0.8929 \\
+        UNet & 7.8 & \textbf{0.8929} \\
+        \hline\hline
         UL-VM-UNet & 0.049 & 0.79955 \\
-        UL-VM-UNet\_v1 & 0.20 & 0.8409 \\
-        UL-VM-UNet\_v2 & 0.042 & 0.8243 \\
-        UL-VM-UNet\_v3 & 0.176 & 0.8109 \\
-        UL-VM-UNet\_v4 & 1.1 & 0.8548 \\
-        UL-VM-UNet\_v5 & 0.7 & 0.8633 \\
+        UL-VM-UNet-v1(Ours) & 0.20 & 0.8409 \\
+        UL-VM-UNet-v2(Ours) & \textbf{0.042} & 0.8243 \\
+        UL-VM-UNet-v3(Ours) & 0.176 & 0.8109 \\
+        UL-VM-UNet-v4(Ours) & 1.1 & 0.8548 \\
+        UL-VM-UNet-v5(Ours) & 0.7 & \textbf{0.8633} \\
+        UL-VM-UNet-v6(Ours) & 0.77 & 0.8628 \\
         \hline
@@
 
-\begin{figure}[!t]
+\begin{figure}[ht]
     \centering
@@
 \end{figure}
-\gr{do we need a table of the params and scores? May be helpful.. Maybe stick one in the appendix. }
 
-\begin{figure}[!t]
+\begin{figure}[ht]
     \centering
-    \includegraphics[width=\columnwidth]{imgs/masks.png}
-    \caption{Example segmentations.}
+    \includegraphics[width=\columnwidth]{imgs/tumor_seg_ex.png}
+    \caption{Sample of segmentation results from each model and their dice score for the segmentation. (Additional samples in Appendix \ref{supp_figs})}
     \label{fig:segs}
@@
 
-\gr{Would appendix really need to be in the page limit? Appendix is to include figures that wouldn't fit in paper?}
+The integrated pyramidal pooling module to UMambaBot resulted in a slight increase in performance over the base model.
+However, UMambaBot-PP had an increase in parameter count by approximately 200,000 parameters.
+Additionally, our UL-VM-UNet-v2 achieved the lowest parameter count of 0.042 M with an increase over the original UltraLight VM-UNet architecture.
+Integrating a pyramidal pooling module to the UltraLight model, as done in UL-VM-UNet-v4 and -v6, resulted in an increase in parameter count with no change or a decrease in performance.
 
@@
 Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
-The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
+The smallest model, ULVM-Net-v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
 
-While the traditional U-Net model achieved the highest dice score of 0.8929, our UMambaBot\_PP model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}.
-Notably, our Mamba-based UMambaBot\_PPBot and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
+While the traditional U-Net model achieved the highest dice score of 0.8929, our UMambaBot-PP model closely followed with a dice score of 0.8867.
+Notably, our Mamba-based UMambaBot-PP and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.
 
-We were able to make the ultra-light model smaller and more accurate on our task, further proving the potential of Mamba-based systems that prioritize compactness and efficiency.
+We were able to make the UltraLight model smaller and more accurate on our task, further proving the potential of Mamba-based systems that prioritize compactness and efficiency.
 The selective state spaces and parallel processing strategies employed in our models enabled us to strike a balance between model size and segmentation accuracy.
@@
 Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.
+
+\bibliographystyle{ieeetr}
+\bibliography{refs}
+
 \appendix
@@
 
-3. Training: The model is trained using the training set for [X] epochs with a batch size of [Y]. We use the [Z] optimizer with a learning rate of [L].
+3. Training: The model is trained using the training set for 100 epochs with a batch size of 32. We use the Adam optimizer with a learning rate of 0.0003.
 
-4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use metrics such as accuracy, precision, recall, and F1-score.
+4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use the dice loss metric.
 
@@
 % Second appendix section
-\subsection{Supplementary Figures}
+\subsection{Supplementary Figures} \label{supp_figs}
 
-\begin{thebibliography}{00}
-    \bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/14_9.png}
+\end{figure*}
 
-    \bibitem{vaswani2023attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2023.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/22_4.png}
+\end{figure*}
 
-    \bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/13_4.png}
+    \caption{In some cases, such as this one, it was observed that models with too few parameters (UltraLight-VM-UNet, UltraLight-VM-UNet-v2) don't detect sections of the tumor during segmentation.}
+\end{figure*}
 
-    \bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/19_4.png}
+    \caption{UltraLight-VM-UNet-v5 achieved segmentation results equivalent to UMambaBot with approximately 9.1 M less parameters.}
+\end{figure*}
 
-    \bibitem{bakas2019identifying} S.
-Bakas et al., "Identifying the Best Machine Learning Algorithms for Brain Tumor Segmentation, Progression Assessment, and Overall Survival Prediction in the BRATS Challenge," arXiv:1811.02629 [cs.CV], 2019.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/0_0.png}
+    \caption{An example of models with pyramidal pooling in the PVM blocks (UltraLight-VM-UNet-v4, UltraLight-VM-UNet-v6) achieved an improvement over a model without (UltraLight-VM-UNet-v5).}
+\end{figure*}
 
-    \bibitem{U-Mamba} J. Ma, F. Li, and B. Wang, "U-Mamba: Enhancing Long-range Dependency for Biomedical Image Segmentation," arXiv:2401.04722 [cs.CV], 2024.
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/12_2.png}
+\end{figure*}
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/22_3.png}
+\end{figure*}
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{imgs/22_7.png}
+\end{figure*}
 
-    \bibitem{rombach2021highresolution} R. Rombach, A. Blattmann, D. Lorenz, P. Esser, and B. Ommer, "High-Resolution Image Synthesis with Latent Diffusion Models," arXiv:2112.10752 [cs.CV], 2021.
-
-    \bibitem{DBLP:journals/corr/abs-2006-11239} J.
-Ho, A.
-Jain, and P.
-Abbeel, "Denoising Diffusion Probabilistic Models," CoRR, vol.
-abs/2006.11239, 2020.
-[Online].
-Available: https://arxiv.org/abs/2006.11239
-
-    \bibitem{DBLP:journals/corr/HeZRS15} K.
-He, X.
-Zhang, S.
-Ren, and J.
-Sun, "Deep Residual Learning for Image Recognition," CoRR, vol.
-abs/1512.03385, 2015.
-[Online].
-Available: http://arxiv.org/abs/1512.03385
-
-    \bibitem{zhang2022segvit} B. Zhang et al., "SegViT: Semantic Segmentation with Plain Vision Transformers," arXiv:2210.05844 [cs.CV], 2022.
-
-    \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
-    \bibitem{ultralightvmunet} R.
-Wu, Y.
-Liu, P.
-Liang, and Q.
-Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr.
-20, 2024.
-[Online].
-Available: https://arxiv.org/pdf/2403.20035.pdf
-    \bibitem{zhao2017pyramid}
-    H.
-Zhao, J.
-Shi, X.
-Qi, X.
-Wang and J.
-Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp.
-6230-6239, doi: 10.1109/CVPR.2017.660.
-    \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
-    \bibitem{xie2021segformer} E.
-Xie, W.
-Wang, Z.
-Yu, A.
-Anandkumar, J.
-Alvarez, and P.
-Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
-\end{thebibliography}
 \end{document}

### 2024-04-26 paper/main.tex
finalized

@@
 
-Making use of Structured State Spaces, presented by Gu, Goel, and Ré \cite{gu2022efficiently}, for their superiority in long range dependency tasks, a specific type of SSM, the Mamba architecture, by Gu and Dao \cite{gu2023mamba}, proposed a foundational model to operate on arbitrary sequences from a variety of inputs in the domain of sequence modeling.
+Making use of Structured State Spaces, presented by Gu, Goel, and Ré \cite{gu2022efficiently}, for their capability in long range dependency tasks, a specific type of SSM, the Mamba architecture, by Gu and Dao \cite{gu2023mamba}, proposed a foundational model to operate on arbitrary sequences from a variety of inputs in the domain of sequence modeling.
 This model achieves the power of Transformers while scaling linearly in sequencing length to be computationally efficient.
@@
 Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models.
-For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{ho2020denoising} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}.
+For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{ho2020denoising} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2022highresolution}.
 We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.
@@
 \subsection{U-Mamba}
-As our model takes much inspiration from the recently developed U-Mamba design \cite{ma2024umamba}, we also choose to incorporate it in our comparison.
+As our models take inspiration from the recently developed U-Mamba design \cite{ma2024umamba}, we also choose to incorporate it in our comparison.
 U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures.
@@
 This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{wu2024ultralight} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.
+The channel list is the same as Ultra LightM-UNet's, [8,16,24,32,48,64].
 
@@
 The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.
+Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+The channel list was [16,32,64,128,256,512].
 
@@
 UL-VM-UNet-v5 is meant to be a comparison to the -v4 and -v6 UltraLight variations.
-This model trained and performed well with 1.1M parameters.
-Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.
+This model trained and performed well with 0.7M parameters.
+v5 has 8 parallels and modifies the channel list to [16,32,64,128,256,512].
 
@@
 The main goal here, again, was to integrate pyramidal pooling for the benefit of multi-scale feature extraction while keeping a low parameter count.
+The channel list was also [16,32,128,256,512].
 
@@
 
-Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
+Our experiments and comparisons with SOTA models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
 The smallest model, ULVM-Net-v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.
@@
 \renewcommand{\thesection}{\Alph{section}}
+
 % First appendix section
-\subsection{Training Pipeline}
-The training pipeline for our model consists of the following steps:
-
-1. Data preprocessing: The dataset is split into training, validation, and test sets. The images are resized to a uniform size and normalized.
-
-2. Model architecture: We employ a convolutional neural network (CNN) with the following layers: [Describe the layers and their configurations].
-
-3. Training: The model is trained using the training set for 100 epochs with a batch size of 32. We use the Adam optimizer with a learning rate of 0.0003.
-
-4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use the dice loss metric.
-
-5. Testing: Once the model achieves satisfactory performance on the validation set, it is tested on the held-out test set to assess its generalization ability.
+\subsection{Training}
+We train multiple models for image segmentation using an 80-10-10 train-test-evaluation split.
+The accuracy of the models is evaluated using Dice scores, and the DiceLoss is used as the loss function.
+\\
+PP-UMamba, UMambaBot, ResNet18, ResNet50, and UNet were trained for 100 epochs using DiceLoss as the loss function and Adam as the optimizer with a learning rate of 3e-4.\\
+UltraLight-VM-UNet and UltraLight-VM-UNet\_v1-v6 were trained for 100 epochs using BCEDiceLoss as the loss function and AdamW as the optimizer with a learning rate of 3e-4.
+All UltraLight variants (v1-v6) are trained using the same settings as UltraLight-VM-UNet.
+\\
+SegFormer was trained for 100 epochs using BCEWithLogitsLoss as the loss function and AdamW as the optimizer with a learning rate of 1e-4.
+The learning rate is reduced on a plateau using a scheduler, and early stopping is employed.
+\\
+More detailed training information can be found in our GitHub repository: \url{https://github.com/lxaw/mamba-tumor-seg/} and under the \texttt{tests} folder.
 

### 2024-04-26 final_upload/tests/calculate_dice_score.py

@@ -1,16 +1,0 @@
-import torch
-
-def calculate_dice_score(predicted_mask, target_mask):
-    # # Flatten predicted and target masks
-    # predicted_flat = predicted_mask.view(-1)
-    # target_flat = target_mask.view(-1)
-
-    # Calculate intersection
-    intersection = torch.sum(predicted_mask * target_mask)
-
-    # Calculate Dice coefficient
-    dice = (2. * intersection) / (torch.sum(predicted_mask) + torch.sum(target_mask))
-
-    
-
-    return dice

### 2024-04-26 final_upload/tests/calculate_iou.py

@@ -1,9 +1,0 @@
-import torch 
-
-def calculate_iou(pred_mask, true_mask):
-
-    intersection = torch.logical_and(pred_mask, true_mask).sum()
-    union = torch.logical_or(pred_mask, true_mask).sum()
-
-    iou = intersection / union if union != 0 else 0.0
-    return iou

### 2024-04-26 final_upload/tests/dice_loss.py

@@ -1,22 +1,0 @@
-import torch
-import torch.nn as nn
-
-class DiceLoss(nn.Module):
-    def __init__(self):
-        super(DiceLoss, self).__init__()
-
-    def forward(self, logits, targets):
-        smooth = 100
-        num = targets.size(0)
-
-        # Apply sigmoid to each tensor in the list
-        probs = [torch.sigmoid(logit) for logit in logits]
-
-        # Stack the tensors along a new dimension
-        probs = torch.stack(probs)
-
-        m1 = probs.view(num, -1)
-        m2 = targets.view(num, -1)
-        intersection = (m1 * m2).sum()
-
-        return 1 - ((2. * intersection + smooth) / (m1.sum() + m2.sum() + smooth))

### 2024-04-26 final_upload/tests/mamba_model.py

@@ -1,248 +1,0 @@
-import torch
-from torch import nn
-from mamba_ssm import Mamba
-
-import os
-from glob import glob
-import pandas as pd 
-import cv2
-import torch
-import torchvision
-from torchvision.models import vgg16_bn
-import torchvision.transforms as transforms
-from sklearn.model_selection import train_test_split
-import torch.nn as nn
-import numpy as np
-import matplotlib.pyplot as plt
-import numpy as np
-import torch
-from torch import nn
-from typing import Union, Type, List, Tuple
-
-from dynamic_network_architectures.building_blocks.helper import get_matching_convtransp
-from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
-from dynamic_network_architectures.building_blocks.plain_conv_encoder import PlainConvEncoder
-from dynamic_network_architectures.building_blocks.residual import StackedResidualBlocks
-
-from dynamic_network_architectures.building_blocks.residual_encoders import ResidualEncoder
-from dynamic_network_architectures.building_blocks.residual import BasicBlockD, BottleneckD
-from torch.nn.modules.conv import _ConvNd
-from torch.nn.modules.dropout import _DropoutNd
-from dynamic_network_architectures.building_blocks.helper import convert_conv_op_to_dim
-
-from nnunetv2.utilities.plans_handling.plans_handler import ConfigurationManager, PlansManager
-from dynamic_network_architectures.building_blocks.helper import get_matching_instancenorm, convert_dim_to_conv_op
-from nnunetv2.utilities.network_initialization import InitWeights_He
-from mamba_ssm import Mamba
-
-
-## Custom class imports
-#
-
-
-from pyramidal_pooling import PyramidalPoolingModule
-
-# Ignore Warnings
-import warnings
-warnings.filterwarnings("ignore")
-
-
-class UNetResDecoder(nn.Module):
-    def __init__(self,
-                 encoder: Union[PlainConvEncoder, ResidualEncoder],
-                 num_classes: int,
-                 n_conv_per_stage: Union[int, Tuple[int, ...], List[int]]):
-        """
-        This class requires the skip connections from the encoder as input during its forward pass.
-
-        The encoder progresses to the bottleneck, where the decoder takes over. The decoder stages are organized based on computation order, with the initial stage having the lowest resolution. In this stage, the decoder takes the bottleneck features and the lowest-resolution skip connection as inputs.
-
-        Each stage of the decoder comprises three parts:
-
[…]

### 2024-04-26 final_upload/tests/transforms.py

@@ -1,17 +1,0 @@
-##################################
-# Clay Crews, Lex Whalen
-#
-# Dataset class for Brain MRI Transforms.
-# For use in later pipelines.
-
-from torchvision import transforms
-
-class BrainMRITransforms:
-    def __init__(self, resize_shape=(256, 256)):
-        self.transforms = transforms.Compose([
-            transforms.ToTensor(),
-            transforms.Resize(resize_shape)
-        ])
-
-    def __call__(self, img):
-        return self.transforms(img)

### 2024-04-26 paper/main.tex
naming

@@
 \\
-PP-UMamba, UMambaBot, ResNet18, ResNet50, and UNet were trained for 100 epochs using DiceLoss as the loss function and Adam as the optimizer with a learning rate of 3e-4.\\
-UltraLight-VM-UNet and UltraLight-VM-UNet\_v1-v6 were trained for 100 epochs using BCEDiceLoss as the loss function and AdamW as the optimizer with a learning rate of 3e-4.
-All UltraLight variants (v1-v6) are trained using the same settings as UltraLight-VM-UNet.
-\\
+UMambaBot-PP, UMambaBot, ResNet18, ResNet50, and UNet were trained for 100 epochs using DiceLoss as the loss function and Adam as the optimizer with a learning rate of 3e-4.\\
+UL-VM-UNet and UL-VM-UNet-v1 through v6 were trained for 100 epochs using BCEDiceLoss as the loss function and AdamW as the optimizer with a learning rate of 3e-4.  \\
 SegFormer was trained for 100 epochs using BCEWithLogitsLoss as the loss function and AdamW as the optimizer with a learning rate of 1e-4.

### 2024-04-29 paper/main.tex
final edits

--- 2024-04-25 paper/main.tex
@@
 
-\section{Proposed Methodology}
-In this paper, we strive to maintain high accuracy in our image segmentation while also keeping low parameter counts.
-We primarily seek to use pyramidal pooling modules \cite{zhao2017pyramid} to improve accuracy, and use a variant of the Mamba \cite{gu2023mamba} architecture inspired by \cite{wu2024ultralight} to maintain small parameter sizes.
-
-\section{Input Data}\label{sec2}
-We shall be using a Brain MRI segmentation dataset found in \cite{dataset}.
-This dataset has been used in several papers regarding classification of the shape and severity of tumors, and provides an adequate dataset for research purposes in the field of medical image analysis, particularly in the area of brain MRI segmentation.
-Researchers have utilized this dataset in various studies focusing on the classification of tumor shapes and the assessment of tumor severity.
-With its comprehensive collection of brain MRI scans, along with corresponding segmentation masks, the dataset offers valuable resources for developing and evaluating algorithms aimed at automating tumor detection and analysis.
-
-The dataset contains brain MRI images along with their segmentation masks for 110 patients.
-There were a total of 3143 train images, 393 validation images, and 393 test images.
-The distribution of the tumor and non-tumor images is shown in Figure \ref{fig:datastats}.
-
-\section{Output Data}\label{sec3}
-The output of this model shall be segmentation masks for new, never before seen images.
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
-    \caption{Distributions of tumor and non-tumor images in the train, validation, and test datasets.}
-    \label{fig:datastats}
-\end{figure}
-
 \section{Mamba}
@@
 
-\section{Compared Models}
+\section{Related Works}
 Prior to an analysis of how the Pyramidal U-Mamba compares, we shall explain what models we chose for comparison and why.
@@
 
+\section{Proposed Methodology}
+In this paper, we strive to maintain high accuracy in our image segmentation while also keeping low parameter counts.
+We primarily seek to use pyramidal pooling modules \cite{zhao2017pyramid} to improve accuracy, and use a variant of the Mamba \cite{gu2023mamba} architecture inspired by \cite{wu2024ultralight} to maintain small parameter sizes.
+
+\section{Input Data}\label{sec2}
+We shall be using a Brain MRI segmentation dataset found in \cite{dataset}.
+This dataset has been used in several papers regarding classification of the shape and severity of tumors, and provides an adequate dataset for research purposes in the field of medical image analysis, particularly in the area of brain MRI segmentation.
+Researchers have utilized this dataset in various studies focusing on the classification of tumor shapes and the assessment of tumor severity.
+With its comprehensive collection of brain MRI scans, along with corresponding segmentation masks, the dataset offers valuable resources for developing and evaluating algorithms aimed at automating tumor detection and analysis.
+
+The dataset contains brain MRI images along with their segmentation masks for 110 patients.
+There were a total of 3143 train images, 393 validation images, and 393 test images.
+The distribution of the tumor and non-tumor images is shown in Figure \ref{fig:datastats}.
+
+\section{Output Data}\label{sec3}
+The output of this model shall be segmentation masks for new, never before seen images.
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
+    \caption{Distributions of tumor and non-tumor images in the train, validation, and test datasets.}
+    \label{fig:datastats}
+\end{figure}
+
 \section{Our Models}
@@
 Model results are shown in Table \ref{tab:model_comparison} and Figures \ref{fig:balls},\ref{fig:segs}, with additional segmentation results in Appendix \ref{supp_figs}.
+Additional training information can be found in Appendix \ref{trainingpipeline}.
 
@@
 % First appendix section
-\subsection{Training Pipeline}
+\subsection{Training Pipeline} \label{trainingpipeline}
 The training pipeline for our model consists of the following steps:

### 2024-04-29 paper/main.tex
final

@@
     Our methodology involves implementing and fine-tuning a Mamba-based architecture on a dataset of annotated brain MRI scans.
-We will evaluate the model's performance using standard segmentation metrics and compare the results with U-net and transformer-based approaches.
+We will evaluate the model's performance using standard segmentation metrics and compare the results with state-of-the-art (SOTA) segmentation approaches.
 We open-source the code here: \href{https://github.com/lxaw/mamba-tumor-seg}{https://github.com/lxaw/mamba-tumor-seg}.
@@
 
-Our goal was to 1) maintain dice scores comparable to state-of-the-art (SOTA) methods such as those listed above, while 2) being smaller than those above.
+Our goal was to 1) maintain dice scores comparable to SOTA methods such as those listed above, while 2) being smaller than those above.
 
@@
 
-Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
+Our experiments and comparisons with state-of-the-art models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy.
 The smallest model, ULVM-Net-v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.

### 2024-04-29 final_upload/tests/brain_mri_dataset.py
delete not needed

@@ -1,54 +1,0 @@
-##################################
-# Clay Crews, Lex Whalen
-#
-# Dataset class for Brain MRI
-# For use in later pipelines.
-
-# Necessary imports
-import torch
-from sklearn.model_selection import train_test_split
-import pandas as pd
-from glob import glob
-import cv2
-import numpy as np
-
-class BrainMRIDatasetBuilder:
-    def __init__(self, data_dir):
-        self.data_dir = data_dir
-
-    def create_df(self):
-        images_paths = []
-        masks_paths = glob(f'{self.data_dir}/*/*_mask*')
-
-        for i in masks_paths:
-            images_paths.append(i.replace('_mask', ''))
-
-        df = pd.DataFrame(data={'images_paths': images_paths, 'masks_paths': masks_paths})
-
-        return df
-
-    def split_df(self, df):
-        train_df, dummy_df = train_test_split(df, train_size=0.8, random_state=37)
-        valid_df, test_df = train_test_split(dummy_df, train_size=0.5, random_state=37)
-        return train_df, valid_df, test_df
-
-
-class BrainMRIDataset(torch.utils.data.Dataset):
-    def __init__(self, dataframe, transform=None, mask_transform=None):
-        self.df = dataframe
-        self.transform = transform
-        self.mask_transform = mask_transform
-
-    def __len__(self):
-        return len(self.df)
-
-    def __getitem__(self, idx):
-        image = cv2.imread(self.df.iloc[idx, 0]) / 255.0
-        mask = cv2.imread(self.df.iloc[idx, 1], cv2.IMREAD_GRAYSCALE) / 255.0
-        mask = np.where(mask >= 0.5, 1., 0.)
-        if self.transform:
-            image = self.transform(image)
-        if self.mask_transform:
-            mask = self.mask_transform(mask)
-
-        return image, mask

### 2024-04-29 final_upload/tests/mamba_pipeline.py

@@ -1,144 +1,0 @@
-################################
-#
-# Clay Crews, Lex Whalen
-# Pyramidal U-Mamba
-# 
-# Explanation: 
-# We take the original U-Mamba architecture (see here: https://github.com/bowang-lab/U-Mamba) and add
-# Pyramidal Pooling 
-#
-import torch
-import torch.nn as nn
-from torch.utils.data import Dataset, DataLoader
-
-from mamba_model import UMambaBot
-from dice_loss import DiceLoss
-import torch.optim as optim
-from brain_mri_dataset import BrainMRIDatasetBuilder,BrainMRIDataset
-
-from transforms import BrainMRITransforms
-
-from calculate_iou import calculate_iou
-
-
-device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
-model = UMambaBot(
-    input_channels=3,  # Assuming RGB images with 3 channels
-    n_stages=5,
-    features_per_stage=(32, 64, 128, 256,512),
-    conv_op=nn.Conv2d,  # Assuming 2D convolution
-    kernel_sizes=(3, 3, 3, 3, 3),  # Adjusted kernel sizes for 2D convolution
-    strides=(1, 2, 2, 2, 2),
-    num_classes=3,
-    n_conv_per_stage=(1, 1, 1, 1, 1),
-    n_conv_per_stage_decoder=(1, 1, 1, 1),
-    conv_bias=True,
-    norm_op=nn.InstanceNorm2d,  # Assuming 2D instance normalization
-    norm_op_kwargs={},
-    dropout_op=None,
-    nonlin=nn.LeakyReLU,
-    nonlin_kwargs={'inplace': True},
-    # Pyramidal Pooling
-    ppm_pool_sizes=(1,2,3,6)
-).to(device)
-
-total_params = sum(p.numel() for p in model.parameters())
-trainable_params = sum(p.numel() for p in model.parameters() if p.requires_grad)
-print(f'Total parameters: {total_params}')
-print(f'Trainable parameters: {trainable_params}')
-
-learning_rate = 0.0003
-
-criterion = DiceLoss()
-optimizer = optim.Adam(
-    model.parameters(),
-    lr=learning_rate)
-
-epochs = 100
-
-train_loss = []
-val_loss = []
-trainIOU = []
-valIOU = []
-
-import sys
-
-builder = BrainMRIDatasetBuilder("../datasets/tumor_segs")
-df = builder.create_df()
[…]

### 2024-04-29 final_upload/tests/mamba_pipeline_test.py

@@ -1,31 +1,0 @@
-import torch
-import torch.nn as nn
-
-from mamba_model import UMambaBot
-
-weights = "mamba_pipeline_py_weights.pth"
-
-device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
-
-model = UMambaBot(
-    input_channels=3,  # Assuming RGB images with 3 channels
-    n_stages=5,
-    features_per_stage=(32, 64, 128, 256,512),
-    conv_op=nn.Conv2d,  # Assuming 2D convolution
-    kernel_sizes=(3, 3, 3, 3, 3),  # Adjusted kernel sizes for 2D convolution
-    strides=(1, 2, 2, 2, 2),
-    num_classes=3,
-    n_conv_per_stage=(1, 1, 1, 1, 1),
-    n_conv_per_stage_decoder=(1, 1, 1, 1),
-    conv_bias=True,
-    norm_op=nn.InstanceNorm2d,  # Assuming 2D instance normalization
-    norm_op_kwargs={},
-    dropout_op=None,
-    nonlin=nn.LeakyReLU,
-    nonlin_kwargs={'inplace': True},
-    # Pyramidal Pooling
-    ppm_pool_sizes=(1,2,3,6)
-).to(device)
-
-model.load_state_dict(torch.load(weights))
-model.eval()

### 2024-04-29 final_upload/tests/pyramidal_pooling.py

@@ -1,34 +1,0 @@
-import torch
-import torch.nn as nn
-
-import torch.nn.functional as F
-
-class PyramidalPoolingModule(nn.Module):
-    def __init__(self, input_channels,pool_sizes,output_channels):
-        super(PyramidalPoolingModule, self).__init__()
-        self.pooling_layers = nn.ModuleList([
-            nn.AdaptiveAvgPool2d((pool_size, pool_size))  # AdaptiveAvgPool2d with variable pool sizes
-            for pool_size in pool_sizes
-        ])
-        if isinstance(output_channels, int):
-            self.output_channels = [output_channels] * len(self.pooling_layers)
-        else:
-            assert len(output_channels) == len(self.pooling_layers), "Length of output_channels must match the number of pooling layers"
-            self.output_channels = output_channels
-        self.conv_layers = nn.ModuleList([
-            nn.Conv2d(input_channels, out_channels, kernel_size=1)  # Convolutional layer to adjust channel dimension
-            for out_channels in self.output_channels
-        ])
-
-    def forward(self, x):
-        pooled_features = []
-        for pooling_layer, conv_layer in zip(self.pooling_layers, self.conv_layers):
-            pooled_feature = pooling_layer(x)
-            pooled_feature = F.relu(conv_layer(pooled_feature))  # Apply convolution to adjust channel dimension
-            pooled_feature = F.interpolate(pooled_feature, size=x.size()[2:], mode='bilinear', align_corners=True)  # Resize to original spatial resolution
-            pooled_features.append(pooled_feature)
-                # Adjust the number of output channels to match the desired final output
-        total_output_channels = sum(self.output_channels)
-        output_features = torch.cat(pooled_features, dim=1)  # Concatenate pooled features along channel dimension
-        output_features = output_features[:, :total_output_channels, :, :]  # Trim to desired output channels
-        return output_features

### 2024-04-29 models/ultralightVMUNet_v6.py
updated ULv6 code

@@ -1,0 +1,286 @@
+import torch
+from torch import nn
+import torch.nn.functional as F
+
+from timm.models.layers import trunc_normal_
+import math
+from mamba_ssm import Mamba
+
+from pyramidal_pooling import PyramidalPoolingModule
+
+
+def load_model_and_weights(device):
+    model = nn.DataParallel(UltraLight_VM_UNet(
+        num_classes=1,
+        input_channels=3,
+        # c_list=[8,16,24,32,48,64],
+        c_list=[16,32,64,128,256,512],
+        split_att='fc',
+        bridge=True
+    )).to(device)
+    # Load weights for parallel training
+    state_dict = torch.load('modified_ultralightmunet_v6_weights.pth', map_location=device)
+    # Remove the 'module.' prefix from keys if present
+    # state_dict = {k.replace('module.', ''): v for k, v in state_dict.items()}
+    # Load state dict
+    model.load_state_dict(state_dict)
+    model.eval()
+
+    return model
+
+
+
+class PVMLayer(nn.Module):
+    def __init__(self, input_dim, output_dim, d_state = 16, d_conv = 4, expand = 2, pp=False):
+        super().__init__()
+        self.input_dim = input_dim
+        self.output_dim = output_dim
+        self.norm = nn.LayerNorm(input_dim)
+        self.mamba = Mamba(
+                d_model=input_dim//8, # Model dimension d_model
+                d_state=d_state,  # SSM state expansion factor
+                d_conv=d_conv,    # Local convolution width
+                expand=expand,    # Block expansion factor
+        )
+        self.proj = nn.Linear(input_dim, output_dim)
+        self.skip_scale= nn.Parameter(torch.ones(1))
+        self.pp = pp
+        
+        if self.pp == True:
+            self.ppm = PyramidalPoolingModule(input_channels=input_dim,
+                                          pool_sizes=(1,2,3,6),
+                                          output_channels=[input_dim//8, input_dim//4, input_dim//2, input_dim//8]) 
+    
+    def forward(self, x):
[…]

### 2024-05-16 paper/main.tex

@@
 Model results are shown in Table \ref{tab:model_comparison} and Figures \ref{fig:balls},\ref{fig:segs}, with additional segmentation results in Appendix \ref{supp_figs}.
-Additional training information can be found in Appendix \ref{trainingpipeline}.
+Additional training information can be found in Appendix \ref{training}.
 
@@
         UMambaBot-PP(Ours) & 10.0 & 0.8867 \\
-        UMambaBot & \textbf{9.8} & 0.8799 \\
+        UMambaBot & 9.8 & 0.8799 \\
         ResNet18 & 13.9 & 0.8648 \\
@@
         SegFormer & 17.8 & 0.8454 \\
-        UNet & 7.8 & \textbf{0.8929} \\
+        UNet & \textbf{7.8} & \textbf{0.8929} \\
         \hline\hline
@@
 % First appendix section
-\subsection{Training Pipeline} \label{trainingpipeline}
-The training pipeline for our model consists of the following steps:
+\subsection{Training Pipeline} \label{training}
 
-1. Data preprocessing: The dataset is split into training, validation, and test sets. The images are resized to a uniform size and normalized.
+We train multiple models for image segmentation using an 80-10-10 train-test-evaluation split. The accuracy of the models is evaluated using Dice scores, and the DiceLoss is used as the loss function.
 
-2. Model architecture: We employ a convolutional neural network (CNN) with the following layers: [Describe the layers and their configurations].
+UMambaBot-PP, UMambaBot, ResNet18, ResNet50, and UNet were trained for 100 epochs using DiceLoss as the loss function and Adam as the optimizer with a learning rate of 3e-4.
 
-3. Training: The model is trained using the training set for 100 epochs with a batch size of 32. We use the Adam optimizer with a learning rate of 0.0003.
+UL-VM-UNet and UL-VM-UNet-v1 through v6 were trained for 100 epochs using BCEDiceLoss as the loss function and AdamW as the optimizer with a learning rate of 3e-4.
 
-4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use the dice loss metric.
+SegFormer was trained for 100 epochs using BCEWithLogitsLoss as the loss function and AdamW as the optimizer with a learning rate of 1e-4.
+The learning rate is reduced on a plateau using a scheduler, and early stopping is employed.
 
-5. Testing: Once the model achieves satisfactory performance on the validation set, it is tested on the held-out test set to assess its generalization ability.
+More detailed training information can be found in our GitHub repository: \url{https://github.com/lxaw/mamba-tumor-seg/} and under the \texttt{tests} folder.
 
